Simple repression draft

### 2020-06-05 doc/section_02_means.tex
Finish appendix on mean and Fanos for nonbursty models
Not quite final pass on everything, Appdx S3 still needs more

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no 
transcription for many of these states. The models are probabilistic with each
state assigned some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key axes: what states to
consider and how to compute the probabilities of those states.

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel away. In addition, the presence of different
operator sites, i.e. binding sites for the repressor, is known to change the
basal expression level in the absence of repressor~\mmnote{look up
cite}\mrm{does reg-seq mention anything like this in their SI? I definitely saw
that on my own experiments, but didn't mention it in the paper}. Computing
fold-changes neatly cancels this effect as well.

@@

\subsubsection{The Two-State Equilibrium Model}
In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
the promoter is idealized as existing in one of two states, either repressor
bound or repressor unbound. The rate of transcription, and therefore the level
of gene expression, is assumed to be proportional to the fraction of time spent
in the repressor unbound state. Mathematically, this amounts to the statement
that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome where repressors spend most
of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
The derivation of this result is deferred to Appendix~\ref{sec:non_bursty}.

The fold-change is  found as the ratio of mean mRNA with and without repressor
as introduced in eqn.~\ref{eq:fc_def}. Invoking that definition results in
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
which clearly matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(C) with $\rho=1$ and $\Delta F_R = 
\beta\Delta\varepsilon_r - \log (R / N_{NS})$.

In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
as the coarse-graining of any equilibrium promoter model in which no
transcriptionally active states have transcription factor bound. Put
differently, there is no overlap between TF bound states and transcriptionally
active states. We will see this explicitly in the 3-state equilibrium model
below, but perhaps surprising is that an analogous result carries over even to
the nonequilibrium models we consider later.

@@

Let $p_{m,R}(t)$ denote the joint probability of finding the system in the
repressor bound state $R$ with $m$ mRNA molecules present at time $t$. Similarly
define $p_{m,U}(t)$ for the repressor unbound state $U$. This model is governed
by coupled master equations giving the time evolution of $p_{m,R}(t)$ and
$p_{m,U}(t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
\begin{align}
\begin{split}
\deriv{t}p_{m,R}(t) =& 
- \overbrace{k_R^- p_{m,R}(t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_{m,U}(t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_{m+1,R}(t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma p_{m,R}(t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_{m,U}(t) =&\; 
\overbrace{k_R^- p_{m,R}(t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_{m,U}(t)}^{U \rightarrow R}
+ \overbrace{rp_{m-1,U}(t)}^{m-1 \rightarrow m}
- \overbrace{rp_{m,U}(t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_{m+1,U}(t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma p_{m,U}(t)}^{m \rightarrow m - 1},
\label{eq:poisson_promoter_cme}
\end{split}
\end{align}
where each term on the right corresponds to a transition between two states of
the promoter as indicated by the overbrace label. In each equation, the first
two terms describe transitions between promoter states due to repressors
unbinding and binding, respectively. The final two terms describe degradation of
mRNA, decreasing the copy number by one, and the terms with coefficient $r$
describe transcription initiation increasing the mRNA copy number by one.

We can greatly simplify the notation, which will be especially useful for the
more complicated models yet to come, by re-expressing the master equation in
vector form~\cite{Phillips2012}. The promoter states are collected into a vector
and the rate constants are collected into matrices as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,U} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\matr{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R} \vect{p}_{m-1}(t)
                + \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:2state_rep_cme}
\end{equation}
where $\matr{I}$ is the identity matrix. Taking steady state by setting time
derivatives to zero, the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
with the algebra details again deferred to Appendix~\ref{sec:non_bursty}. Recall
$k_R^+$ is proportional to the repressor copy number, so in computing
fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.

@@

The master equation governing this model is analogous
to~\eqrange{eq:2state_cme_matrices}{eq:2state_rep_cme} for model 1 above. The
main subtlety arises since transcription initiation accompanies a promoter state
change. This can be understood by analogy to $\matr{K}$. The off-diagonal and
diagonal elements of $\matr{K}$ correspond to transitions arriving at or
departing from, respectively, the promoter state of interest. If transcription
initiation is accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since the arriving and
departing transitions have different initial copy numbers of mRNA, unlike for
$\matr{K}$ where they are the same (see Appendix~\cite{sec:non_bursty}). The
master equation for this model is
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\ p_{m,P} \end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vect{p}_m$ encode the probabilities of having $m$ mRNA present
along with the promoter having repressor bound ($R$), being empty ($E$), or
having RNAP bound ($P$), respectively. $\matr{R_A}$ describes probability flux
\textit{arriving} at the state $\vect{p}_m$ from a state with one fewer mRNA,
namely $\vect{p}_{m-1}$, and $\matr{R_D}$ describes probability flux
\textit{departing} from the state $\vect{p}_m$ for a state with one more mRNA,
namely $\vect{p}_{m+1}$. $\matr{K}$ is closely analogous to model 1.

Mean mRNA at steady state is found analogously to model 1, with the result
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}.
\end{equation}
Fold-change is again found from the ratio prescribed by~\eq{eq:fc_def}, from
which we have
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
        \frac{k_P^+ + k_P^- + r}{k_P^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
        \right)^{-1}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
        \right)^{-1},
\end{align}
which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
k_P^+/(k_P^- + r)$ as claimed.

@@

The master equation governing this model is identical in form to model 2 above,
namely
\begin{equation}
\deriv{t}\vect{p}_m(t) =
\left( \matr{K} - \matr{R_D} - \gamma m \matr{I} \right) \vect{p}_m(t)
                + \matr{R_A} \vect{p}_{m-1}(t) +
                \gamma (m+1) \matr{I} \vect{p}_{m+1}(t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices. The
state vector and promoter transition matrix are now
\begin{equation}
\vect{p}_m = \begin{pmatrix} p_{m,R} \\ p_{m,E} \\
                             p_{m,C} \\ p_{m,O}\end{pmatrix},\
\matr{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound ($R$), empty
($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
increasing dimension by one, the only new feature in $\matr{K}$ is the
rate $k_O$, representing the (irreversible) rate of open complex formation from
the closed complex. The initiation matrices are given by
\begin{equation}
\matr{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0
                \end{pmatrix},\
\matr{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to nonequilibrium model 2.

The expression for mean mRNA is substantially more complicated now, as worked
out in Appendix~\ref{sec:non_bursty} where we find
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+ k_O}
        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
\end{equation}
which can be simplified to
\begin{equation}
\langle m\rangle
= \frac{r}{\gamma}
\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
\end{equation}
The strategy is to isolate the terms involving the repressor, so that now the
fold-change is seen to be simply
\begin{align}
\text{fold-change}
&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
\\
&= \left(
        1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
\right)^{-1},
\end{align}
surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.

This example hints that an arbitrarily fine-grained model of downstream
transcription steps may still be collapsed to the form of the master curve for
the means giving in~\fig{fig1:means_cartoons}(D) so long as the repressor
binding is exclusive with transcriptionally active states. We offer this as a
conjecture, and we suspect that a careful argument using the King-Altman diagram
method~\cite{King1956, Hill1966} might furnish a ``proof.'' Our focus here is
not on full generality but rather to survey an assortment of plausible models
for simple repression  that have been proposed in the literature.

@@

Mathematically this model is very similar to models 1 and 2. Like model 1, the
matrix $R$ describing transcription initiation is diagonal, namely
\begin{equation}
\matr{R} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
        \end{pmatrix}.
\end{equation}
The master equation takes verbatim the same form as it did for model
1,~\eq{eq:2state_rep_cme}. Meanwhile the promoter transition matrix $K$ is the
same as~\eq{eq:3state_cme_matrices_pt1} from model 2, although we relabel the
rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
RNAP binding and unbinding rates.

@@

\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
intuitive analog to model 1, with just two states, repressor bound or unbound,
and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
the unbound state, single mRNA transcripts are produced as a Poisson process
with some characteristic rate $r$. The current model by contrast produces, at
some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts with a
geometrically distributed burst size and a mean burst size $b$.

@@

Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all?
\textcolor{blue}{This brings to mind the deep questions posed in,
e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, asking us to search for general
``symmetries'' of the problem instead of specifying generative models of the
known microscopics of a system} \mrm{I don't understand this last sentence. Can
we make it more transparent?}. In our case, the key feature seems to be the
exclusive nature of repressor and RNAP binding, which allows the parameter
describing the repressor, $\Delta F_R$, to cleanly separate from all other
details of the promoter architecture, which are encapsulated in $\rho$. In the
graph-theoretic language which was applied to similar problems
in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the repressor-bound state is
a leaf of the graph describing transitions between promoter states. Arbitrary
nonequilibrium behavior can occur on the rest of the promoter state space, but
it may all be swept up in the effective parameter $\rho$, to which the repressor
makes no contribution.

>>>>

As noted in the previous section, there are two broad classes of models in play
for computing the input-output functions of regulatory architectures as shown in
Figure~\ref{fig1:means_cartoons}. In both classes of model, the promoter is
imagined to exist in a discrete set of states of occupancy, with each such state
of occupancy accorded its own rate of transcription -- including no 
transcription for many of these states. The models are probabilistic with each
state assigned some probability and the overall rate of transcription given by 
\begin{equation}
\mbox{average rate of transcription} = \sum_i r_i p_i,
\label{eq:transcrip_prop_pbound}
\end{equation}
where $i$ labels the distinct states, $p_i$ is the probability of the
$i^{\text{th}}$ state, and $r_i$ is the rate of transcription of that state.
Ultimately, the different models differ along several key axes: what states to
consider and how to compute the probabilities of those states.

@@

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Brewster2014, Razo-Mejia2018, Chure2019}. Part of its
utility in dissecting transcriptional regulation is its ratiometric nature,
which removes many secondary effects that are present when making an absolute
gene expression measurement. Also, by measuring otherwise identical cells with
and without a transcription factor present, any biological noise common to both
conditions can be made to cancel away.
% \mmnote{I'd love to include the factoid below but I can't find
% a cite to back it up...}
% In addition, the presence of different
% operator sites, i.e. binding sites for the repressor, is known to change the
% basal expression level in the absence of repressor~\mmnote{look up
% cite}\mrm{does reg-seq mention anything like this in their SI? I definitely saw
% that on my own experiments, but didn't mention it in the paper}. Computing
% fold-changes neatly cancels this effect as well.

@@

\subsubsection{The Two-State Equilibrium Model}
In this simplest model, depicted as (1) in Figure~\ref{fig1:means_cartoons}(B),
the promoter is idealized as existing in one of two states, either repressor
bound or repressor unbound. The rate of transcription
is assumed to be proportional to the fraction of time spent
in the repressor unbound state.
From the relative statistical weights listed in
Figure~\ref{fig1:means_cartoons}, the probability $p_U$ of being in the
unbound state is
\begin{equation}
p_U = \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1}.
\end{equation}
The mean rate of transcription is then given by $r p_U$, as assumed by
Eq.~\ref{eq:transcrip_prop_pbound}. The mean number of mRNA is
set by the balance of average mRNA transcription and degradation rates,
so it follows that the mean mRNA level is given by
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $r$ is the transcription rate from the repressor unbound state, $\gamma$
is the mRNA degradation rate, $R$ is repressor copy number, $N_{NS}$ is the
number of nonspecific binding sites in the genome where repressors spend most
of their time when not bound to the operator, $\beta \equiv 1/k_BT$, and
$\Delta\varepsilon_R$ is the binding energy of a repressor to its operator site.
The derivation of this result is deferred to Appendix~\ref{sec:non_bursty}.

The fold-change is  found as the ratio of mean mRNA with and without repressor
as introduced in Eq.~\ref{eq:fc_def}. Invoking that definition results in
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
which clearly matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R = 
\beta\Delta\varepsilon_r - \log (R / N_{NS})$.

In fact it was noted in~\cite{Chure2019} that this two-state model can be viewed
as the coarse-graining of any equilibrium promoter model in which no
transcriptionally active states have transcription factor bound, or put
differently, when there is no overlap between
transcription factor bound states and transcriptionally
active states. We will see this explicitly in the 3-state equilibrium model
below, but perhaps surprising is that an analogous result carries over even to
the nonequilibrium models we consider later.

@@

Let $p_R(m,t)$ denote the joint probability of finding the system in the
repressor bound state $R$ with $m$ mRNA molecules present at time $t$. Similarly
define $p_U(m,t)$ for the repressor unbound state $U$. This model is governed
by coupled master equations giving the time evolution of $p_R(m,t)$ and
$p_U(m,t)$~\cite{Sanchez2008, Sanchez2011, Phillips2019} which we can write as
\begin{align}
\begin{split}
\deriv{t}p_R(m,t) =& 
- \overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
+ \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{(m+1)\gamma p_R(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma p_R(m,t)}^{m \rightarrow m - 1}
\\
\deriv{t}p_U(m,t) =&\; 
\overbrace{k_R^- p_R(m,t)}^{R \rightarrow U}
- \overbrace{k_R^+ p_U(m,t)}^{U \rightarrow R}
+ \overbrace{rp_U(m-1,t)}^{m-1 \rightarrow m}
- \overbrace{rp_U(m,t)}^{m \rightarrow m + 1}
\\
&+ \overbrace{(m+1)\gamma p_U(m+1,t)}^{m + 1 \rightarrow m}
- \overbrace{\gamma p_U(m,t)}^{m \rightarrow m - 1},
\label{eq:poisson_promoter_cme}
\end{split}
\end{align}
where each term on the right corresponds to a transition between two states of
the promoter as indicated by the overbrace label. In each equation, the first
two terms describe transitions between promoter states due to repressors
unbinding and binding, respectively. The final two terms describe degradation of
mRNA, decreasing the copy number by one, and the terms with coefficient $r$
describe transcription initiation increasing the mRNA copy number by one.
We direct the reader to Appendix~\ref{sec:cme_from_cartoon} for a careful
treatment showing how the form of this master equation follows from the
corresponding cartoon in Figure~\ref{fig1:means_cartoons}.

We can greatly simplify the notation, which will be especially useful for the
more complicated models yet to come, by re-expressing the master equation in
vector form~\cite{Phillips2012}. The promoter states are collected into a vector
and the rate constants are collected into matrices as
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_U(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ \\ k_R^- & -k_R^+ \end{pmatrix},\
\mathbf{R} = \begin{pmatrix} 0 & 0 \\ 0 & r \end{pmatrix},\
\label{eq:2state_cme_matrices}
\end{equation}
so that the master equation may be condensed as
\begin{equation}
\deriv{t}\vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R} \vec{p}(m-1,t)
                + \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:2state_rep_cme}
\end{equation}
where $\mathbf{I}$ is the identity matrix. Taking steady state by setting time
derivatives to zero, the mean mRNA can be found to be
\begin{equation}
\langle m \rangle = \frac{r}{\gamma}
        \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\label{eq:mean_m_model1}
\end{equation}
with the algebra details again deferred to Appendix~\ref{sec:non_bursty}. Recall
$k_R^+$ is proportional to the repressor copy number, so in computing
fold-change, absence of repressor corresponds to $k_R^+\rightarrow0$. Therefore
fold-change in this model is simply
\begin{equation}
\text{fold-change} = \left(1 + \frac{k_R^+}{k_R^-}\right)^{-1},
\end{equation}
again matching the master curve of~\fig{fig1:means_cartoons}(D) with $\rho=1$.

@@

The master equation governing this model is analogous
to Eqs.~\ref{eq:2state_cme_matrices}-\ref{eq:2state_rep_cme} for model 1 above. The
main subtlety arises since transcription initiation accompanies a promoter state
change. This can be understood by analogy to $\mathbf{K}$. The off-diagonal and
diagonal elements of $\mathbf{K}$ correspond to transitions arriving at or
departing from, respectively, the promoter state of interest. If transcription
initiation is accompanied by promoter state changes, we must have separate
matrices for arriving and departing transcription events since the arriving and
departing transitions have different initial copy numbers of mRNA, unlike for
$\mathbf{K}$ where they are the same (see Appendix~\ref{sec:non_bursty}). The
master equation for this model is
\begin{equation}
\deriv{t}\vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\label{eq:3state_rep_cme}
\end{equation}
with the state vector and promoter transition matrix defined as
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\ p_P(m) \end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 \\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- \\
                        0 & k_P^+ & -k_P^- 
                \end{pmatrix},
\label{eq:3state_cme_matrices_pt1}
\end{equation}
and the initiation matrices given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & r \\ 
                0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
                \end{pmatrix}.
\label{eq:3state_cme_matrices_pt2}
\end{equation}
The elements of $\vec{p}(m)$ encode the probabilities of having $m$ mRNA present
along with the promoter having repressor bound ($R$), being empty ($E$), or
having RNAP bound ($P$), respectively. $\mathbf{R_A}$ describes probability flux
\textit{arriving} at the state $\vec{p}(m)$ from a state with one fewer mRNA,
namely $\vec{p}(m-1)$, and $\mathbf{R_D}$ describes probability flux
\textit{departing} from the state $\vec{p}(m)$ for a state with one more mRNA,
namely $\vec{p}(m+1)$. $\mathbf{K}$ is closely analogous to model 1.

Mean mRNA at steady state is found analogously to model 1, with the result
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)},
\label{eq:model2_meanm}
\end{equation}
and with details again deferred to Appendix~\ref{sec:non_bursty}.
Fold-change is again found from the ratio prescribed by Eq.~\ref{eq:fc_def}, from
which we have
\begin{align}
\text{fold-change}
&=      \frac{k_R^- k_P^+}
        {k_R^- k_P^+ + k_R^- (k_P^- + r) + k_R^+ (k_P^- + r)}
        \frac{k_P^+ + k_P^- + r}{k_P^+}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
                \frac{k_P^- + r}{k_P^+ + k_P^- + r}
        \right)^{-1}
\\
&=      \left(1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+}{k_P^- + r}\right)^{-1}
        \right)^{-1},
\end{align}
which follows the master curve of~\fig{fig1:means_cartoons}(D) with $\rho = 1 +
k_P^+/(k_P^- + r)$ as claimed.

@@

The master equation governing this model is identical in form to model 2 above,
namely
\begin{equation}
\deriv{t}\vec{p}(m,t) =
\left( \mathbf{K} - \mathbf{R_D} - \gamma m \mathbf{I} \right) \vec{p}(m,t)
                + \mathbf{R_A} \vec{p}(m-1,t) +
                \gamma (m+1) \mathbf{I} \vec{p}(m+1,t),
\end{equation}
but obviously with a higher-dimensional state space and different matrices. The
state vector and promoter transition matrix are now
\begin{equation}
\vec{p}(m) = \begin{pmatrix} p_R(m) \\ p_E(m) \\
                             p_C(m) \\ p_O(m)\end{pmatrix},\
\mathbf{K} = \begin{pmatrix} -k_R^- & k_R^+ & 0 & 0\\
                        k_R^- & -k_R^+ -k_P^+ & k_P^- & 0 \\
                        0 & k_P^+ & -k_P^- - k_O & 0 \\
                        0 & 0 & k_O & 0
                \end{pmatrix},
\end{equation}
with the four promoter states, in order, being repressor bound ($R$), empty
($E$), RNAP closed complex ($C$), and RNAP open complex ($O$). Besides
increasing dimension by one, the only new feature in $\mathbf{K}$ is the
rate $k_O$, representing the (irreversible) rate of open complex formation from
the closed complex. The initiation matrices are given by
\begin{equation}
\mathbf{R_A} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0
                \end{pmatrix},\
\mathbf{R_D} = \begin{pmatrix}
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & 0 \\ 
        0 & 0 & 0 & r
                \end{pmatrix},
\end{equation}
again closely analogous to nonequilibrium model 2.

The expression for mean mRNA is substantially more complicated now, as worked
out in Appendix~\ref{sec:non_bursty} where we find
\begin{equation}
\langle m\rangle = \frac{r}{\gamma}
        \frac{k_R^- k_P^+ k_O}
        {k_R^- [(k_P^+ (k_O + r) + r(k_P^- + k_O)] + k_R^+ r(k_P^- + k_O)},
\label{eq:model3_mean_m}
\end{equation}
which can be simplified to
\begin{equation}
\langle m\rangle
= \frac{r}{\gamma}
\frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}.
\end{equation}
The strategy is to isolate the terms involving the repressor, so that now the
fold-change is seen to be simply
\begin{align}
\text{fold-change}
&= \frac{\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
        {1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)} + \frac{k_R^+}{k_R^-}}
        \frac{1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}}
                {\frac{k_P^+ k_O}{r(k_O + k_P^-)}}
\\
&= \left(
        1 + \frac{k_R^+}{k_R^-}
        \left(1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}\right)^{-1}
\right)^{-1},
\end{align}
surprisingly reducing to the master curve of~\fig{fig1:means_cartoons}(D) once
again, with $\rho = 1 + \frac{k_P^+ (k_O + r)}{r(k_O + k_P^-)}$.

This example hints that an arbitrarily fine-grained model of downstream
transcription steps may still be collapsed to the form of the master curve for
the means given in Figure~\ref{fig1:means_cartoons}(D), so long as the repressor
binding is exclusive with transcriptionally active states. We offer this as a
conjecture, and we suspect that a careful argument using the King-Altman diagram
method~\cite{King1956, Hill1966} might furnish a ``proof.'' Our focus here is
not on full generality but rather to survey an assortment of plausible models
for simple repression  that have been proposed in the literature.

@@

Mathematically this model is very similar to models 1 and 2. Like model 1, the
matrix $R$ describing transcription initiation is diagonal, namely
\begin{equation}
\mathbf{R} = \begin{pmatrix}
                0 & 0 & 0 \\ 
                0 & 0 & 0 \\ 
                0 & 0 & r
        \end{pmatrix}.
\end{equation}
The master equation takes verbatim the same form as it did for model 1,
Eq.~\ref{eq:2state_rep_cme}. Meanwhile the promoter transition
matrix $\mathbf{K}$ is the same as Eq.~\ref{eq:3state_cme_matrices_pt1}
from model 2, although we relabel the
rate constants from $k_P^\pm$ to $k^\pm$ to reiterate that these are not simply
RNAP binding and unbinding rates.

@@

\subsubsection{Nonequilibrium Model Five - Bursty Promoter}
The final model we consider shown in Figure~\ref{fig1:means_cartoons}(C) is an
intuitive analog to model 1, with just two states, repressor bound or unbound,
and transition rates between them of $k_R^+$ and $k_R^-$. In model 1, when in
the unbound state, single mRNA transcripts are produced as a Poisson process
with some characteristic rate $r$. The current model by contrast produces, at
some Poisson rate $k_i$, \textit{bursts} of mRNA transcripts.
The burst sizes are assumed to be geometrically distributed with
a mean burst size $b$.

@@

Why does the master curve of~\fig{fig1:means_cartoons}(D) exist at all?
This brings to mind the deep questions posed in,
e.g.,~\cite{Frank2013} and~\cite{Frank2014a}, suggesting we consider multiple
plausible models of a system and search for their common patterns to tease
out which broad features are and are not important.
In our case, the key feature seems to be the
exclusive nature of repressor and RNAP binding, which allows the parameter
describing the repressor, $\Delta F_R$, to cleanly separate from all other
details of the promoter architecture, which are encapsulated in $\rho$. In the
graph-theoretic language which was applied to similar problems
in~\cite{Gunawardena2012} and~\cite{Ahsendorf2014}, the repressor-bound state is
a leaf of the graph describing transitions between promoter states. Arbitrary
nonequilibrium behavior can occur on the rest of the promoter state space, but
it may all be swept up in the effective parameter $\rho$, to which the repressor
makes no contribution.
